#extras
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# section on plotting data and reconstructions on trees (largely in phytools)

## Plotting reconstructions of traits

Let's have a look at plotting trees and plotting data on them. Because it's cool and I can, I'm going to use a dinosaur phylogeny.
```{r}
theropods <- read.tree("Theropod trees Hedman Benson 2017.tre")[[1]]
theropods <- compute.brlen(theropods, method = "Grafen")
```

\begin{center}
\includegraphics[height=5in,width=5in]{~/Google Drive/University of Liverpool/2018-19/PhyloGroup/Workshops/Workshop 4/Files/TheropodaInfobox.png}
\end{center}

\newpage

Here's the basic tree with some options for customisation.
```{r, fig.height=7.5, fig.width=6}
plot(theropods, 
     cex = 0.5,                     #Tip label font size
     no.margin = T,                 #Get rid of the margin around the plot
     label.offset = 0.01,           #Set the tips away from the labels
     edge.width = 2,                #Thickness of the branches
     edge.color = "skyblue",        #Colour of the branches
     tip.color = "red")             #Colour of the tip labels
```

Liam Revell's package Phytools is great for plotting data along with phylogenies. We've used the package throughout these tutorial so you should be pretty familiar with it by now.

First I'm going to simulate some data.
```{r}
x <- fastBM(theropods)
```

This first plot is a phenogram and shows the trait of interest and how it changes over time with the tree. As you can see it's not terribly clear but depending on your trait, it can reveal some interesting groupings.
```{r, fig.height=5, fig.width=6}
phenogram(theropods, x, fsize = 0.7, ftype = "i")
```

\newpage

As the data we've simulated in continuous, we can plot it as a cont map. This plot mapps the change onto the branches of the phylogeny itself.
```{r, fig.height=8.5, fig.width=6}
obj <- contMap(theropods, x, fsize = 0.6, lwd = 2, outline = F)
```

\newpage

If this doesn't tickle your fancy, you can plot the data on the tips as bars. This commapnd creates a hybrid of a tree and a barplot.
```{r, fig.height=7, fig.width=6}
plotTree.wBars(theropods, x, scale = 0.04)
```

\newpage

If that looks a little odd, remember that our data was simulated either side of 0 and so has positive and negative values. Real data might look more like this.
```{r, fig.height=7, fig.width=6}
y <-fastBM(theropods, bounds = c(0,1))
plotTree.wBars(theropods, y, scale = 0.04)
```

\newpage

With a bit more experience, you can combine some features!
```{r, fig.height=7, fig.width=6}
obj <- contMap(theropods, y, plot = F)
plotTree.wBars(obj$tree, y,
               method = "plotSimmap",
               colors = obj$cols,
               scale = 0.04)
add.color.bar(.5, obj$cols, 
              title = "trait value",
              lims = obj$lims,
              prompt = F,
              x = -1*par()$usr[1],
              y = -32*par()$usr[3])
```

\newpage

For binary character states, we can plot a density map.
```{r}
Q <- matrix(c(-1,1,1,-1),2,2)
rownames(Q) <- colnames(Q) <- c(0,1)
tree <- sim.history(theropods, Q)
z <- tree$states
```

```{r}
trees <- make.simmap(tree, z, nsim = 100)
```

```{r, fig.height=8.5, fig.width=6}
obj <- densityMap(trees, lwd = 3, fsize = 0.6, outline = F)
```

