#extras
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# section on plotting data and reconstructions on trees (largely in phytools)

## Plotting reconstructions of traits

Let's have a look at plotting trees and plotting data on them. Because it's cool and I can, I'm going to use a dinosaur phylogeny.
```{r}
theropods <- read.tree("Theropod trees Hedman Benson 2017.tre")[[1]]
theropods <- compute.brlen(theropods, method = "Grafen")
```

\begin{center}
\includegraphics[height=5in,width=5in]{~/Google Drive/University of Liverpool/2018-19/PhyloGroup/Workshops/Workshop 4/Files/TheropodaInfobox.png}
\end{center}

\newpage

Here's the basic tree with some options for customisation.
```{r, fig.height=7.5, fig.width=6}
plot(theropods, 
     cex = 0.5,                     #Tip label font size
     no.margin = T,                 #Get rid of the margin around the plot
     label.offset = 0.01,           #Set the tips away from the labels
     edge.width = 2,                #Thickness of the branches
     edge.color = "skyblue",        #Colour of the branches
     tip.color = "red")             #Colour of the tip labels
```

Liam Revell's package Phytools is great for plotting data along with phylogenies. We've used the package throughout these tutorial so you should be pretty familiar with it by now.

First I'm going to simulate some data.
```{r}
x <- fastBM(theropods)
```

This first plot is a phenogram and shows the trait of interest and how it changes over time with the tree. As you can see it's not terribly clear but depending on your trait, it can reveal some interesting groupings.
```{r, fig.height=5, fig.width=6}
phenogram(theropods, x, fsize = 0.7, ftype = "i")
```

\newpage

As the data we've simulated in continuous, we can plot it as a cont map. This plot mapps the change onto the branches of the phylogeny itself.
```{r, fig.height=8.5, fig.width=6}
obj <- contMap(theropods, x, fsize = 0.6, lwd = 2, outline = F)
```

\newpage

If this doesn't tickle your fancy, you can plot the data on the tips as bars. This commapnd creates a hybrid of a tree and a barplot.
```{r, fig.height=7, fig.width=6}
plotTree.wBars(theropods, x, scale = 0.04)
```

\newpage

If that looks a little odd, remember that our data was simulated either side of 0 and so has positive and negative values. Real data might look more like this.
```{r, fig.height=7, fig.width=6}
y <-fastBM(theropods, bounds = c(0,1))
plotTree.wBars(theropods, y, scale = 0.04)
```

\newpage

With a bit more experience, you can combine some features!
```{r, fig.height=7, fig.width=6}
obj <- contMap(theropods, y, plot = F)
plotTree.wBars(obj$tree, y,
               method = "plotSimmap",
               colors = obj$cols,
               scale = 0.04)
add.color.bar(.5, obj$cols, 
              title = "trait value",
              lims = obj$lims,
              prompt = F,
              x = -1*par()$usr[1],
              y = -32*par()$usr[3])
```

\newpage

For binary character states, we can plot a density map.
```{r}
Q <- matrix(c(-1,1,1,-1),2,2)
rownames(Q) <- colnames(Q) <- c(0,1)
tree <- sim.history(theropods, Q)
z <- tree$states
```

```{r}
trees <- make.simmap(tree, z, nsim = 100)
```

```{r, fig.height=8.5, fig.width=6}
obj <- densityMap(trees, lwd = 3, fsize = 0.6, outline = F)
```



## Plotting with ggplot2

A nice way of investigating data is by plotting it. For a quick visualisation I usually use the base graphics in R. They get the job done and are realtively simple to edit once you understand the syntax of R.

However, most R users seem to agree that the package "ggplot2" gives better plots. This might be useful to you when you want to prepare your plots for reports. So here, I'm going to use ggplot2 just to show you what it can do.
```{r, eval = FALSE}
install.packages("ggplot2")
library(ggplot2)
```

ggplot2 builds plots in layers similar to ggtree which you have met before in workshop 1. In fact, the ggtree package was built based upon ggplot2. We start with the function ggplot which creates a coordinate system to which we can add layers. This function alone just creates a blank plot
```{r, fig.align='center'}
ggplot(data = primate.data)
```

Let's add a layer with the points. We can add a layer of points with the function geom_point. This creates our basic scatterplot. Each geom function (the ones that add layers) takes a mapping argument which controls how the layer is mapped onto the plot. This argument usually takes the form seen below. Remember that body mass should be log transformed before we plot it!
```{r, fig.align='center'}
ggplot(data = primate.data) +
  geom_point(mapping = aes(x = log10(AdultBodyMass_g), y = GestationLen_d))
```

In practice, exactly how you add geoms and layers seems to be affected by exactly what you want to do with the plot. As we'll see later, if you want to add regression lines or other statistical features to your plot, you may want to reorganise this code. My advice is to google what you want to do and work backwards from someone else's code. That's what I do.

The plot isn't done yet. We can display a bit more information by adding another mapping argument. Here, I've coloured the points according to their family. The unique colours are assigned auotmatically and the legend is also created automatically.
```{r, fig.align='center', fig.height = 4.5, fig.width = 7}
ggplot(data = primate.data) +
  geom_point(mapping = aes(x = log10(AdultBodyMass_g), y = GestationLen_d, 
                           colour = Family))
```

If you prefer, you can manually set the colour for all points. Be aware of the difference here! In the code below, the difference is that colour here is outside of the mapping argument. This means that the colour doesn't display any real information but does change the aesthetic of the plot.
```{r, fig.align='center'}
ggplot(data = primate.data) +
  geom_point(mapping = aes(x = log10(AdultBodyMass_g), y = GestationLen_d),
             colour = "green")
```

Let's clean up the plot with some axis labels.
```{r, fig.align='center', fig.height = 4.5, fig.width = 7}
ggplot(data = primate.data) +
  geom_point(mapping = aes(x = log10(AdultBodyMass_g), y = GestationLen_d,
                           colour = Family)) +
  labs(x = "Log Body Mass", y = "Gestation Length")
```

